---
title: "nodbi-overview"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{nodbi-overview}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
  
```{r, include = FALSE}
knitr::opts_chunk$set(
collapse = TRUE,
comment = "#>"
)
```

# Overview

`nodbi` is an R package that provides a single interface for several NoSQL databases and databases with JSON functionality, with the same function parameters and return values across all six databases: 

- MongoDB 
- SQLite 
- PostgreSQL 
- DuckDB 
- Elasticsearch 
- CouchDB 

```{r setup}
library(nodbi)
```

## Functionality

### Connect

With any of the six databases, `nodbi` functions work in the same way and return the same values. 

First, a connection to a database as `src` is established. In the example, no additional parameters were used such as file or server, see the help pages for the respective database. 

"Container" is used as term to indicate where conceptually the database holds the data. The `key` parameter holds the name of a container. 

```{r}
# name of container
key <- "my_container"

# nodbi can connect any of these databases
if (FALSE) {
  src <- src_duckdb()
  src <- src_mongo(collection = key)
  src <- src_sqlite()
  src <- src_postgres()
  src <- src_elastic()
  src <- src_couchdb(
    user = Sys.getenv("COUCHDB_TEST_USER"), 
    pwd = Sys.getenv("COUCHDB_TEST_PWD"))
}

# this example is run with
src <- src_sqlite()

# note additional parameters can be specified,
# for example for local or remote MongoDb:
# help("src_mongo")
```

### docdb_create

Check and create container. The return value is the number of created documents. Documents refers to the rows in a data frame such as `mtcars`, or the number of `NDJSON` lines, or the number of list items, or the number of objects in an `JSON` array. 

```{r}
# check if container already exists
docdb_exists(src = src, key = key)

# load data (here data frame, alternatively a list, JSON or file with NSJSON)
# into the container "my_container" specified in "key" parameter
docdb_create(src = src, key = key, value = mtcars)

# load additionally 98 NDJSON records
docdb_create(src, key, "https://httpbin.org/stream/98")

# load additionally mapdata as list
docdb_create(src, key, jsonlite::fromJSON(mapdata, simplifyVector = FALSE))

# show JSON structure of contacts
jsonlite::minify(contacts) 

# load additionally contacts JSON data
docdb_create(src, key, contacts)
```

Check and list any other containers exist in the database:

```{r}
docdb_list(src = src)
```

### Identifiers

The unique document identifier is its `_id`. The `_id`'s of an input `value` are either the row names of a data frame (such as `mtcars`) or top-level elements with the name `_id` such as in `contacts` as shown just above. 

Thus, expect a warning when trying to create documents with `_id`'s that already exist. 

The return value can be `0` when no documents could newly be created, or the number of the subset of documents in `value` that did not yet exist and could newly be created. (For updating existing documents, see below function `docdb_update()`.) 

```{r}
# zero new documents created
docdb_create(src, key, value = mtcars)
```

### docdb_get

All documents in a container can now be retrieved with `docdb_get`. 

```{r}
# load library for more 
# readable print output
library(tibble)

# get all documents, irrespective of schema
as_tibble(docdb_get(src, key))

# get just 2 documents using limit and note that
# only fields for these documents are returned
as_tibble(docdb_get(src, key, limit = 2L))
```

### docdb_query

One of the most powerful functions of `nodbi` is `docdb_query` because it permits to combine a query to select documents and a filter to select fields of interest. 

Both parameters `query` (obligatory) and `fields` (optional) use, across all databases, MongoDB syntax such as documented for [queries](https://www.mongodb.com/docs/manual/crud/) and [fields](https://www.mongodb.com/docs/manual/tutorial/project-fields-from-query-results/). 

```{r}
# query for some documents
docdb_query(src = src, key = key, query = '{"mpg": {"$gte": 30}}')
```

Unless `fields` specifies `"_id": 0`, the `_id` field is always included in the output of `docdb_query`.

```{r}
# query some fields from some documents; 'query' is a mandatory 
# parameter and is used here in its position in the signature
docdb_query(src, key, '{"mpg": {"$gte": 30}}', fields = '{"wt": 1, "mpg": 1}')
```


```{r}
# query some fields from some documents, limit return to one document
docdb_query(src, key, '{"mpg": {"$gte": 30}}', fields = '{"_id": 0, "mpg": 1}', limit = 1L)
```

Queries can be more complex such as in this example, showing a dot notation of a sub-field and an example operator (regular expression). 

```{r}
# query some subitem fields from some documents
str(docdb_query(
  src, key, 
  query = '{"$or": [{"age": {"$gt": 21}}, 
           {"friends.name": {"$regex": "^B[a-z]{3,9}.*"}}]}', 
  fields = '{"age": 1, "friends.name": 1}'))
```

Queries work across documents of different structure such as here. 

```{r}
# query with results across documents
docdb_query(
  src, key, 
  query = '{"$or": [{"age": {"$gt": 21}}, {"mpg": {"$gte": 30}}]}', 
  fields = '{"name": 1, "disp": 1}')
```

### Field names

The `JSON` data handle by package `nodbi` may have a large number of field included nested fields in objects (see for example `name` within array `friends` above), and there is no simple function to list field names. 

Thus, an argument is provided for `docdb_query` so that the function returns only the comprehensive list of all field names in documents selected with a query (or in all documents in the container if `query = "{}"` is specified). 

```{r}
docdb_query(src, key, query = '{"_id": {"$regex": "^[0-9]"}}', listfields = TRUE)
```

The dot notation is a path from a root field to the nested field, and this notation can be used in `query` and `fields` parameters of `docdb_query`. 

### docdb_update

Queries can also be used for updating (patching) selected documents with a new `value`(s). The return value of `docdb_update` corresponds to the number of documents that were updated. 

This is another powerful function because  `value` can come from a JSON string, a data frame, a list or a file with `NDJSON` lines, and if `value` includes already `_id`'s, these are used to identify the documents that should be updated. The function uses specific and fast `SQL` functions for most databases. 

```{r}
# number of documents corresponding to query
nrow(docdb_query(src, key, query = '{"carb": 3}'))
 
# update all documents using JSON, replacing the previously existing values
docdb_update(src, key, value = '{"vs": 9, "xy": [1, 2]}', query = '{"carb": 3}')

# show updated values 
docdb_query(src, key, query = '{"carb": 3}', fields = '{"xy": 1}')
```

### docdb_delete

Documents and containers can be deleted with `docdb_delete`. Its return value corresponds to the success of the delete operation. 

```{r}
# number of documents corresponding to query
nrow(docdb_query(src, key, query = '{"age": {"$lte": 23}}'))

# to delete selected documents, specify a query parameter
docdb_delete(src, key, query = '{"age": {"$lte": 23}}')

# this deletes the complete container from database
docdb_delete(src, key)

# check if still exists
docdb_exists(src, key)
```

### Disconnect and shutdown

Package `nodbi` provides an automatic mechanism for shutting down, at the time of `quit` or session restart, those databases that require it (SQLite, DuckDB, PostgreSQL). 

Nevertheless, it is good practice to manually disconnect and shut down connections as such as specific to the database, for example for DuckDB:  

```{r}
src

# shutdown
DBI::dbDisconnect(src$con, shutdown = TRUE); rm(src)
```
